---
title: "Birthday Problem Simulation"
author: "Lucas Brizola Fontoura"
date: "08/2025"
output: 
  md_document:
    variant: gfm
enconding: UTF-8
---

# What is the Birthday Problem?
The Birthday Problem is a famous problem in probability theory that asks how likely it is that in a group of people, at least two will share the same birthday.
The problem is counterintuitive because it suggests that a relatively small group of people can have a high probability of sharing a birthday.

### The question is: How many people do you need in a room for there to be a better than even chance that at least two people share a birthday?

```{r}
tries <- 10000
verbose <- FALSE
```

```{r}
# Simulate with 2 people only
same_birthday <- 0
people <- 2

for (i in 1:tries) {

  people_vector <- c()

  for (j in 1:people) {

    birthday <- sample(1:365, 1)
    people_vector <- c(people_vector, birthday)

    if (verbose) {
      cat("Try", i, ":", j, "person's birthday is", birthday, "\n")
    }

  }

  if (length(unique(people_vector)) < length(people_vector)) {
    same_birthday <- same_birthday + 1
  }

}

cat("Days with same birthday:", same_birthday, "out of", tries, "\n")

cat("Probability of two people sharing a birthday",
    same_birthday / tries,
    "(", (same_birthday / tries) * 100, "% chance)\n")

cat("Probability of two people not sharing a birthday",
    1 - (same_birthday / tries),
    "(", (1 - (same_birthday / tries)) * 100, "% chance)\n")
```

```{r}
# Simulate from 2 to 30 people
for (i in 2:30) {

  same_birthday <- 0

  for (j in 1:tries) {

    people_vector <- c()

    for (k in 1:i) {

      birthday <- sample(1:365, 1)
      people_vector <- c(people_vector, birthday)

      if (verbose) {
        cat("Try", j, ":", k, "person's birthday is", birthday, "\n")
      }

    }

    if (length(unique(people_vector)) < length(people_vector)) {
      same_birthday <- same_birthday + 1
    }
  }

  cat("Days with same birthday for", i, "people:",
      same_birthday, "out of", tries, "\n")

  cat("Probability of in", i, "people someone sharing a birthday",
      same_birthday / tries,
      "(", (same_birthday / tries) * 100, "% chance)\n")

  cat("Probability of in", i, "people no one sharing a birthday",
      1 - (same_birthday / tries),
      "(", (1 - (same_birthday / tries)) * 100, "% chance)\n")

  cat("----------------------------------------------------------------\n")
}
```

# Conclusion

The simulation shows that as the number of people in a group increases, the probability of at least two people sharing a birthday also increases significantly.
In fact, with just 23 people, there is over a 50% chance that at least two people share a birthday.
This counterintuitive result highlights the surpring nature of probability and how out intuition can often lead us astray.