---
title: "Monty Hall Problem Simulation"
author: "Lucas Brizola Fontoura"
date: "08/2025"
output: 
  md_document:
    variant: gfm
enconding: UTF-8
---

# What is the Monty Hall Problem?
The Monty Hall problem is a famous probability puzzle based on a game show scenario.
The problem is name after Monty Hall, the original host of the American television game show Let's Make a Deal.
In the game, a contestant is presented with three doors. Behind one door is a car (the prize), and behind the other two doors are goats.
The contestant picks one door, then the host, who knows what is behind each door, opens another door that has a goat behind it.
The contestant is then given the option to either stick with their original choice or switch to the other unopened door.

### The question is: Is it better to switch or to stay with the original choice?

```{r}
tries <- 1000
verbose <- TRUE
```

```{r}
# First strategy: stay with the original choice
correct <- 0
for (i in 1:tries) {
  prize <- sample(1:3, 1) # Randomly place the car behind one of the doors
  choice <- sample(1:3, 1) # Contestant makes an initial choice

  if (verbose) {
    cat("Try", i,
        "- Prize is behind door", prize,
        ", contestant chose door", choice, "\n")
  }

  if (choice == prize) {
    correct <- correct + 1 # Contestant wins if they stay with their choice
  }
}

cat("Staying with the original choice, the contestant wins",
    correct, "out of", tries, "tries.\n")
```

```{r}
# Second strategy: switch to the other door
correct <- 0
for (i in 1:tries) {
  prize <- sample(1:3, 1) # Randomly place the car behind one of the doors
  choice <- sample(1:3, 1) # Contestant makes an initial choice

  open_door <- setdiff(1:3, c(prize, choice)) # Host opens a door with a goat
  switch_choice <-
    setdiff(1:3, c(choice, open_door[1])) # The other unopened door

  if (verbose) {
    cat("Try", i,
        " - Prize is behind door", prize,
        ", contestant chose door", choice,
        ", host opened door", open_door[1],
        ", contestant switches to door", switch_choice, "\n")
  }

  if (switch_choice == prize) {
    correct <- correct + 1 # Contestant wins if they switch to the other door
  }
}

cat("Switching to the other door, the contestant wins",
    correct, "out of", tries, "tries.\n")
```

# Conclusion
The Monty Hall problem demonstrates that switching doors increases the probability of winning from 1/3 to 2/3.
This counterintuitive result highlights the importance of understanding conditional probability and the impact of additional information on decision-making.

### Always switch doors! 